Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-06-30 paper/ifo-introduction.tex
editor comments for abstract, intro and secion ii

In 2015, the first detection of gravitational waves from the merger of two black holes marked a breakthrough in modern astrophysics and revealed a new means to observe the Universe~\cite{GW150914}.
Gravitational waves are a prediction of Albert Einstein's theory of General Relativity~\cite{Einstein:1916}, 
they are disturbances in spacetime caused by the acceleration of asymmetric massive objects.
The effect of gravitational waves is a change in lengths: a ``stretching and squashing'' of the distance between objects. 
Ground-based gravitational-wave observatories such as the Advanced Laser Interferometer Gravitational-wave Observatory (LIGO~\cite{AdvancedLIGO:2015}), Advanced Virgo~\cite{AdvancedVirgo:2015}, GEO600~\cite{GEO600:2010}, and KAGRA~\cite{KAGRA:2013} use the interference of laser light to measure changes in distance. 
These observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top interferometers are commonly used in undergraduate laboratory experiments~\cite{UgoliniEtAl:2019,LibbrechtBlack:2015} and to demonstrate the science of gravitational-wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,AMIGO:online,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

>>>>

In 2015, the first detection of gravitational waves from the merger of two black holes marked a breakthrough in modern astrophysics and revealed a new means to observe the Universe~\cite{GW150914}.
Gravitational waves are a prediction of Albert Einstein's theory of General Relativity~\cite{Einstein:1916}; they are disturbances in spacetime caused by the acceleration of asymmetric massive objects.
The effect of gravitational waves is a change in lengths: a ``stretching and squashing'' of the distance between objects. 
Ground-based gravitational-wave observatories such as the Advanced Laser Interferometer Gravitational-wave Observatory (LIGO~\cite{AdvancedLIGO:2015}), Advanced Virgo~\cite{AdvancedVirgo:2015}, GEO600~\cite{GEO600:2010}, and KAGRA~\cite{KAGRA:2013} use the interference of laser light to measure changes in distance. 
These observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top interferometers are commonly used in undergraduate laboratory experiments~\cite{UgoliniEtAl:2019,LibbrechtBlack:2015} and to demonstrate the science of gravitational-wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,AMIGO:online,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

### 2021-06-30 paper/ifo-tableTopGWs.tex
editor comments for abstract, intro and secion ii

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms, shown in Fig.~\ref{fig:ifo_schematic_webcam}. 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance traveled by the beams. 
Current generation gravitational-wave observatories have kilometer-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo, respectively). 
They can measure minuscule changes in distance due to gravitational waves, e.g.\ the first detection of a binary black hole merger produced a strain of $10^{-21}$~\cite{GW150914}, corresponding to a change in distance which is a fraction of the width of a proton.

Sound is a commonly used analogy when explaining gravitational-wave science. 
Gravitational-wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}.
Detection and analysis techniques used by the gravitational-wave community can be demonstrated using table-top equipment (see Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals.

@@

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a personal computer (PC). 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2020}.
The intensity timeseries is recorded via webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.
In Section~\ref{sec:optical_microphone}, which considers more complex audio, the webcam is replaced with a photodiode, which offers a higher sample rate than the webcam.

@@

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, then the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
But if the relative change in the length of the arms is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency.
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
% too much to have --- and () in the same sentence I think 
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.
We derive this nonlinear relationship between the length difference of the arms and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

>>>>

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms, shown in Fig.~\ref{fig:ifo_schematic_webcam}. 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance traveled by the beams. 
Current generation gravitational-wave observatories have kilometer-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo, respectively). 
They can measure minuscule changes in distance due to gravitational waves; e.g., the first detection of a binary black hole merger produced a strain of $10^{-21}$~\cite{GW150914}, corresponding to a change in distance equal to a fraction of the width of a proton.

Sound is a commonly used analogy when explaining gravitational-wave science. 
Gravitational-wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}.
Detection and analysis techniques used by the gravitational-wave community can be demonstrated using table-top equipment (see Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals.

@@

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a personal computer (PC). 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2020}.
\textbf{The intensity timeseries recorded by the webcam is discussed in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.}
When capturing more complex audio (Section~\ref{sec:optical_microphone}), the webcam is replaced by a photodiode, which offers a higher sample rate than the webcam.

@@

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, then the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
For larger relative length changes, multiple fringes will pass through the detection point during a single speaker oscillation, raising the measured frequency artificially. 
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
% too much to have --- and () in the same sentence I think 
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.
We derive this nonlinear relationship between the length difference of the arms and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

### 2021-06-30 paper/paper_main.tex
editor comments for abstract, intro and secion ii

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational-wave detector: a Michelson interferometer where sound is used as an analog for gravitational waves. 
Using signal processing techniques from continuous-wave searches, we demonstrate the recovery of tones with constant and wandering frequencies. 
We also explore the use of the interferometer as a teaching tool for educators in physics and electrical engineering by using it as an ``optical microphone'' to capture music and speech. 
A range of filtering techniques used to recover signals from noisy data are detailed in the Supplementary Material. 
Here, we present highlights of our results using a combined notch plus Wiener filter and the statistical log minimum mean-square error (logMMSE) estimator. 
Recordings of simple chords and drums are easily recovered using these techniques, but complex music and speech are more challenging. 
This demonstration can be used by educators in undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences.  
\end{abstract}

>>>>

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational-wave detector: a Michelson interferometer where sound is used as an analog for gravitational waves. 
Using signal processing techniques from continuous-wave searches, we demonstrate the recovery of tones with constant and wandering frequencies. 
We also explore the use of the interferometer as a teaching tool for educators in physics and electrical engineering by using it as an ``optical microphone'' to capture music and speech. 
A range of filtering techniques used to recover signals from noisy data are detailed in the Supplementary Material. 
Here, we present highlights of our results using a combined notch plus Wiener filter and the statistical log minimum mean-square error (logMMSE) estimator. 
Recordings of simple chords and drums are easily recovered using these techniques, but complex music and speech are more challenging. 
This demonstration can be used by educators in undergraduate laboratories and can be adapted for communicating gravitational-wave and signal-processing topics to non-specialist audiences.  
\end{abstract}
